---
title: "groupedHG: Hypergeometric & Binomial Group Sampling"
author: "Sumonkanti Das"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{groupedHG: Hypergeometric & Binomial Group Sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(groupedHG)
library(ggplot2)
library(ggpubr)
```

# Introduction

# Introduction

This document introduces the **groupedHG** R package, which provides tools for designing and analyzing group (pooled) sampling under imperfect testing conditions. In many biological and public‐health settings, pooled sampling is used to detect pest species, pathogens, or estimate disease prevalence more efficiently. However, real‐world tests—such as PCR or serological assays—are seldom perfect: sensitivity often declines due to dilution in large pools, and specificity can vary with assay conditions (Arnold *et al.*, 2005). These imperfections can bias prevalence estimates and lead to suboptimal decision‐making, especially during disease outbreaks or biosecurity monitoring (Barnes *et al.*, 2025).

The `groupedHG` package provides an R implementation of the methods developed in Barnes *et al.* (2025), offering exact hypergeometric and fast binomial‐approximation routines that explicitly model sensitivity ($\Delta$, $\delta$) and specificity ($\Lambda$, $\lambda$) at the group or item level. It implements the sampling models—whether detecting any contaminated item or scaling with the number of contaminated items—along with Hellinger information–based tools for optimizing inference accuracy. Unlike existing R packages such as `binGroup` or `pooled`, `groupedHG` brings together the hypergeometric formulas, binomial approximations, and information‐theoretic design principles from Barnes *et al.* into a single, accessible framework.

This vignette is organized as follows:

1. **Mathematical Background & Notation**  
   Definitions of $N$, $T_X$, $b$, $\bar{N}$, and the core hypergeometric/binomial formulas.  
2. **Basic Usage & Case Study**  
   Demonstrations of PMF and Hellinger‐information functions, including a Salmonella surveillance example in pig pens.  
3. **Design Optimization**  
   Grid‐search exploration of pool size ($\bar{N}$) and number of pools ($b$) via Hellinger information heatmaps.  
4. **Session Info & References**  
   `sessionInfo()` output and full bibliographic entries.


# Installation

You can install the development version of `groupedHG` as below.

```r
install.packages("devtools")
devtools::install_github("sumon148/groupedHG")
```

# Mathematical Background

The `groupedHG` package implements statistical models for analyzing group (pooled) sampling data under both perfect and imperfect testing conditions. It is designed to support the inference of contamination prevalence and optimize sampling strategies in contexts such as biosecurity, public health, and ecology. The package provides exact formulas based on the hypergeometric distribution, as well as computationally efficient binomial approximations. It also accounts for test imperfections at both the group level (using parameters $\Delta$ and $\Lambda$) and the item level (using $\delta$ and $\lambda$). Additionally, the package offers tools for computing Hellinger information, which supports the design of statistically efficient group testing schemes.


## Notation

- $N$: population size  
- $T_X$: number of contaminated items  
- $b$: number of pools  
- $\bar{N}$: pool size
- $t_y$: number of positive pools  



## Core Formulas

- `pmf_hg_perfect`: Hypergeometric PMF under perfect testing 
- `pmf_hg_group_imperfect`: Hypergeometric PMF with group-level imperfections ($\Delta / \Lambda$)  
- `pmf_hg_item_imperfect`: Hypergeometric PMF with item-level imperfections ($\delta / \lambda$)  
- `pmf_bn_perfect`: Binomial approximation under perfect testing  
- `pmf_bn_group_imperfect`: Binomial approximation with group-level imperfections ($\Delta / \Lambda$)  
- `pmf_bn_item_imperfect`: Binomial approximation with item-level imperfections ($\delta / \lambda$)  
- `info_hg_tx_imperfect`: Hellinger information for hypergeometric model with imperfections  
- `info_bn_tx_imperfect`: Hellinger information for binomial approximation with imperfections

## Correspondence to Equations in Barnes et al. (2025) paper

The following equations from the Barnes et al. (2025) form the statistical foundation of the `groupedHG` package:

- **Equation (4):** Hypergeometric PMF under perfect testing  
- **Equation (6):** Hypergeometric PMF with group-level imperfections  
- **Equation (8):** Hypergeometric PMF with item-level imperfections  
- **Equation (15):** Approximate binomial PMF under perfect testing  
- **Equation (16):** Binomial approximation with group-level imperfections  
- **Equation (17):** Binomial approximation with item-level imperfections  
- **Equation (20):** Hellinger information computed from the PMF with respect to the considered distribution


# Basic Usage of HG modeling

## PMF under perfect tests 

Consider a study population of $N=100$ items, among which $T_X=20$ are contaminated. These items are divided into $b=4$ groups of equal size. Our goal is to estimate the probability mass function (PMF) for the possible number of detected positive groups, denoted by $t_y$, under the assumption of perfect test sensitivity and specificity. Using the `pmf_hg_perfect()` function, we calculate the PMF values for all possible $t_y$ from 0 to 4. This provides the probabilities of observing each count of detected groups, which can then be used to understand the distribution of group-level test outcomes in the population.

```{r PMF HG Perfect}
b=4
ty.values <- c(0:b)
PRty.HG.perfect <- sapply(ty.values, 
function(ty) pmf_hg_perfect(
ty, N=100, barN=4, Tx=20, b=b))
PRty.HG.perfect
plot(0:b, PRty.HG.perfect, type = "l", col = "black", xlab=expression(paste("Detected groups, ", t[y])),
     ylab = "PMF", main = "Grouped HG",lty=2)
points(0:b,PRty.HG.perfect,type = "b", pch = 19, col = "black")
```

Next, we extend the analysis by examining the PMF functions that account for imperfect testing at both the group and item levels. By applying the pmf_hg_group_imperfect() and pmf_hg_item_imperfect() functions, we can evaluate how test sensitivity and specificity affect the distribution of detected positive groups. We begin by calculating the PMF for the number of detected positive groups assuming a perfect test with 100% sensitivity (`delta=1`) and specificity (`lambda=1`). This serves as a baseline for comparison. 

```{r PMF HG Perfect Group}
PRty.HG.group.perfect <- sapply(ty.values, 
function(ty) pmf_hg_group_imperfect(
ty, N=100, barN=4, Tx=20, b=b, 
delta=1, lambda=1,
verbose = FALSE))
PRty.HG.group.perfect
```

In similar way, the function for item-level imperfect test provides the same results.

```{r PMF HG Perfect Item}
PRty.HG.item.perfect <- sapply(ty.values, 
function(ty) pmf_hg_item_imperfect(
ty, N=100, barN=4, Tx=20, b=b, 
delta=1, lambda=1,
verbose = FALSE))
PRty.HG.item.perfect
```


## PMF under imperfect tests

To investigate the impact of imperfect testing, we compute the PMF using both group-level and item-level sensitivity and specificity. In both cases, we assume a sensitivity of $\Delta=\delta=0.7$ and a specificity of $\Lambda=\lambda=0.8$. Upper and lower case Greek letters refer to respectively group and item level test accuracy. The function `pmf_hg_group_imperfect()` models the scenario where the test responds to the presence or absence of any contaminated item in a group (group-level accuracy), while `pmf_hg_item_imperfect()` accounts for the influence of each individual item's contamination status (item-level accuracy). As a result, the PMF distributions produced by these functions differ: group-level imperfections tend to **underestimate the number of positive groups** when contamination is sparse, while item-level imperfections may result in **greater uncertainty**, especially when multiple contaminated items exist within a group. Comparing these distributions to the perfect test case reveals how diagnostic accuracy directly influences the detection probabilities and, ultimately, prevalence inference.

```{r PMF HG Imperfect Group}
PRty.HG.group.imperfect <- sapply(ty.values, 
function(ty) pmf_hg_group_imperfect(
ty, N=100, barN=4, Tx=20, b=b, 
delta=0.70, lambda=0.80,
verbose = FALSE))
PRty.HG.group.imperfect
```

```{r PMF HG Imperfect Item}
PRty.HG.item.imperfect <- sapply(ty.values, 
function(ty) pmf_hg_item_imperfect(
ty, N=100, barN=4, Tx=20, b=b, 
delta=0.70, lambda=0.80,
verbose = FALSE))
PRty.HG.item.imperfect
```

The difference due to the group- and item-level test accuracy is prortrain in the below figure.

```{r PMF HG Imperfect}
plot(0:b, PRty.HG.item.imperfect, type = "l", col = "red", xlab=expression(paste("Detected groups, ", t[y])),
     ylab = "PMF", main = "Grouped HG",lty=2)
points(0:b,PRty.HG.group.imperfect,type = "b", pch = 19, col = "black")
points(0:b,PRty.HG.item.imperfect,type = "b",lty=2, pch = 19, col = "red")
legend("topleft",legend=c("Group","Item"),col=c(1,2),lty=c(1,2),bty = "n")
```


# Binomial Approximations of HG model

Now we will use binomial approximation of HG model to estimate the probability of having positive groups under perfect and imperfect test. Let use the R functions developed considering test accuracy to estimate the probability of having positive groups $t_y=0,1,2,3,4$. 

```{r PMF BN Perfect}
PRty.BN.perfect <- sapply(ty.values, 
function(ty) pmf_bn_perfect(
ty, N=100, barN=4, Tx=20, b=b))
PRty.BN.group.perfect <- sapply(ty.values, 
function(ty) pmf_bn_group_imperfect(
ty, N=100, barN=4, Tx=20, b=b,delta=1,lambda=1))
PRty.BN.item.perfect <- sapply(ty.values, 
function(ty) pmf_bn_item_imperfect(
ty, N=100, barN=4, Tx=20, b=b,delta=1,lambda=1))
rbind(ty=c(0:b),PRty.BN.perfect,PRty.BN.group.perfect,PRty.BN.item.perfect)
```

Assuming group level test inaccuracy, we wish to estimate the probability and then compare with those assuming item-level test inaccuracy.

```{r PMF BN Imperfect Group}
PRty.BN.group.imperfect <- sapply(ty.values, 
function(ty) pmf_bn_group_imperfect(
ty, N=100, barN=4, Tx=20, b=b, 
delta=0.70, lambda=0.80))
PRty.BN.item.imperfect <- sapply(ty.values, 
function(ty) pmf_bn_item_imperfect(
ty, N=100, barN=4, Tx=20, b=b, 
delta=0.70, lambda=0.80))
plot(0:b, PRty.BN.item.imperfect, type = "l", col = "red", xlab=expression(paste("Detected groups, ", t[y])),
     ylab = "PMF", main = "Grouped BN",lty=2)
points(0:b,PRty.BN.group.imperfect,type = "b", pch = 19, col = "black")
points(0:b,PRty.BN.item.imperfect,type = "b",lty=2, pch = 19, col = "red")
legend("topleft",legend=c("Group","Item"),col=c(1,2),lty=c(1,2),bty = "n")

```

The differences in the probability mass function (PMF) between the exact HG model and its BN approximation can be examined below. We plotted the PMFs under both perfect and imperfect testing scenarios for both HG and BN sampling models. In the context of imperfect testing, group-level test characteristics are denoted by $D_\Delta$ (sensitivity) and $D_\Lambda$ (specificity), while item-level characteristics are denoted by $D_\delta$ and $D_\lambda$, respectively. The plot reveals that the BN approximation closely matches the exact HG model.


```{r PMF HG BN Perfect Imperfect Group Item}
range.pmf <- range(c(PRty.HG.perfect,PRty.BN.perfect,PRty.HG.group.imperfect,PRty.HG.item.imperfect,PRty.BN.group.imperfect,PRty.BN.item.imperfect))
plot(0:b, PRty.HG.perfect, type = "b", lty=2,  pch = 19, col = "black", xlab=expression(paste("Detected groups, ", t[y])),
     ylab = "PMF", main = "Grouped HG / BN",ylim=range.pmf)
points(0:b,PRty.BN.perfect,type = "b", lty=3, pch = 21, col = "black")
points(0:b,PRty.HG.group.imperfect,type = "b",lty=2, pch = 19, col = "red")
points(0:b,PRty.BN.group.imperfect,type = "b",lty=3, pch = 21, col = "red")
points(0:b,PRty.HG.item.imperfect,type = "b", lty=2, pch = 19, col = "blue")
points(0:b,PRty.BN.item.imperfect,type = "b",lty=3, pch = 21, col = "blue")
legend("topleft",
       legend = c(
         "Perfect Case: HG",
         "Perfect Case: BN",
         bquote(HG ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(BN ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(HG ~ Imperfect ~ D[delta] * "," ~ G[lambda]),
         bquote(BN ~ Imperfect ~ D[delta] * "," ~ G[lambda])
       ),
       col = c("black", "black", "red", "red", "blue", "blue"),
       lty = c(2, 3, 2, 3, 2, 3),     # match line types for each model
       pch = c(19, 21, 19, 21, 19, 21), # match point types for each model
       bty = "n", 
       cex = 0.8)
```


# Hellinger Information

Now we work with Hellinger information, a generalized version of Fisher Information, to evaluate the amount of information available for estimating prevalence given a fixed number of contaminated items ($T_X$). This metric quantifies the sensitivity of the PMF to changes in $T_X$, and helps assess the efficiency of different group sampling designs—especially under imperfect testing conditions.

The `info_hg_tx_imperfect()` function computes Hellinger information for the grouped hypergeometric model under imperfect testing conditions. It supports both group- and item-level test characteristics, allowing comparison of different information measures such as analytically driven formula (method="AD"). The below chunk estimates HI for $T_X=0:80$ under HG sampling with perfect and imperfect scenarioes.


```{r HI HG Perfect Imperfect}
Tx.values <- 0:80

# Safe extractor for FI_Tx
extract_FI <- function(x) {
  if (is.list(x) && "FI_Tx" %in% names(x)) {
    return(x$FI_Tx)
  } else {
    return(0)
  }
}

# Perfect test (delta=1, lambda=1)
FI.HG.Perfect <- lapply(
  Tx.values, function(Tx) info_hg_tx_imperfect(
    Tx = Tx, N = 100, b = 4, barN = 4, 
    delta = 1.0, lambda = 1.0, 
    method = "PMF-HI", type = "item", verbose = FALSE
  )
)
FI.Tx.HG.Perfect <- sapply(FI.HG.Perfect, extract_FI)

# Group-level imperfect test
FI.HG.Group.Imperfect <- lapply(
  Tx.values, function(Tx) info_hg_tx_imperfect(
    Tx = Tx, N = 100, b = 4, barN = 4, 
    delta = 0.7, lambda = 0.8, 
    method = "PMF-HI", type = "group", verbose = FALSE
  )
)
FI.Tx.HG.Group.Imperfect <- sapply(FI.HG.Group.Imperfect, extract_FI)

# Item-level imperfect test
FI.HG.Item.Imperfect <- lapply(
  Tx.values, function(Tx) info_hg_tx_imperfect(
    Tx = Tx, N = 100, b = 4, barN = 4, 
    delta = 0.7, lambda = 0.8, 
    method = "PMF-HI", type = "item", verbose = FALSE
  )
)
FI.Tx.HG.Item.Imperfect <- sapply(FI.HG.Item.Imperfect, extract_FI)


```

We now compare the Hellinger information across perfect testing, and group- and item-level imperfect testing scenarios, for the full range of contaminated item counts $T_X = 0$ to $80$, under the hypergeometric (HG) sampling model.


```{r Compare HI HG}
range.HI <- range(c(FI.Tx.HG.Perfect,FI.Tx.HG.Group.Imperfect,FI.Tx.HG.Item.Imperfect))
plot(Tx.values, FI.Tx.HG.Perfect, type = "l", lty=2,  col = "black", xlab=expression(paste("Contaminated items, ", T[X])),
     ylab = "HI of Tx", main = "HI: Grouped HG / BN",ylim=range.HI)
points(Tx.values,FI.Tx.HG.Group.Imperfect,type = "l",lty=3, col = "red")
points(Tx.values,FI.Tx.HG.Item.Imperfect,type = "l", lty=4, col = "red")
legend("topright",
       legend = c(
         "Perfect Case: HG",
         bquote(HG ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(HG ~ Imperfect ~ D[delta] * "," ~ G[lambda])
         ),
       col = c("black", "red", "red"),
       lty = c(2, 3, 4),     # match line types for each model
       bty = "n", 
       cex = 0.8)
```

In the similar way we can estimate HI for the BN sampling model as below.

```{r HI BN Perfect Imperfect}
Tx.values <- c(0:80)
# Different cases: Item level sensitivity
FI.BN.Perfect <- lapply(
  Tx.values, function(Tx) info_bn_tx_imperfect(
    Tx, N=100, b=4, barN=4, delta=1.0, lambda=1.0, 
    method = "PMF-HI", type = "item"))
FI.Tx.BN.Perfect <- unlist(FI.BN.Perfect)
FI.BN.Group.Imperfect <- lapply(
  Tx.values, function(Tx) info_bn_tx_imperfect(
    Tx, N=100, b=4, barN=4, delta=0.7, lambda=0.8, 
    method = "PMF-HI", type = "group"))
FI.Tx.BN.Group.Imperfect <- unlist(FI.BN.Group.Imperfect)
FI.BN.Item.Imperfect <- lapply(
  Tx.values, function(Tx) info_bn_tx_imperfect(
    Tx, N=100, b=4, barN=4, delta=0.7, lambda=0.8, 
    method = "PMF-HI", type = "item"))
FI.Tx.BN.Item.Imperfect <- unlist(FI.BN.Item.Imperfect)
```

Now we will compare the HI for $T_X$ under HG and BN approximation.

```{r HI HG BN Perfect Imperfect Group Item}
par(mfrow=c(1,3))
Tx.values <- c(0:80)
range.HI <- range(c(FI.Tx.HG.Perfect,FI.Tx.BN.Perfect))
plot(Tx.values, FI.Tx.HG.Perfect, type = "l", lty=2,  pch = 19, col = "black", xlab=expression(paste("Detected groups, ", t[y])),
     ylab = "HI of Tx", main = "HI: Grouped HG / BN",ylim=range.HI)
points(Tx.values,FI.Tx.BN.Perfect,type = "l", lty=3, pch = 21, col = "black")
legend("topright",
       legend = c(
         "Perfect Case: HG",
         "Perfect Case: BN",
         bquote(HG ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(BN ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(HG ~ Imperfect ~ D[delta] * "," ~ G[lambda]),
         bquote(BN ~ Imperfect ~ D[delta] * "," ~ G[lambda])
       ),
       col = c("black", "black", "red", "red", "blue", "blue"),
       lty = c(2, 3, 2, 3, 2, 3),     # match line types for each model
       pch = c(19, 21, 19, 21, 19, 21), # match point types for each model
       bty = "n", 
       cex = 0.8)

Tx.values <- c(0:80)
range.HI <- range(c(FI.Tx.HG.Group.Imperfect,FI.Tx.BN.Group.Imperfect))
plot(Tx.values, FI.Tx.HG.Group.Imperfect, type = "l",lty=2, pch = 19, col = "red", xlab=expression(paste("Detected groups, ", t[y])),
     ylab = "HI of Tx", main = "HI: Grouped HG / BN",ylim=range.HI)
points(Tx.values,FI.Tx.BN.Group.Imperfect,type = "l",lty=3, pch = 21, col = "red")
legend("topright",
       legend = c(
         "Perfect Case: HG",
         "Perfect Case: BN",
         bquote(HG ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(BN ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(HG ~ Imperfect ~ D[delta] * "," ~ G[lambda]),
         bquote(BN ~ Imperfect ~ D[delta] * "," ~ G[lambda])
       ),
       col = c("black", "black", "red", "red", "blue", "blue"),
       lty = c(2, 3, 2, 3, 2, 3),     # match line types for each model
       pch = c(19, 21, 19, 21, 19, 21), # match point types for each model
       bty = "n", 
       cex = 0.8)


Tx.values <- c(0:80)
range.HI <- range(c(FI.Tx.HG.Item.Imperfect,FI.Tx.BN.Item.Imperfect))
plot(Tx.values, FI.Tx.HG.Item.Imperfect, type = "l", lty=2, pch = 19, col = "blue", xlab=expression(paste("Detected groups, ", t[y])),
     ylab = "HI of Tx", main = "HI: Grouped HG / BN",ylim=range.HI)
points(Tx.values,FI.Tx.BN.Item.Imperfect,type = "l",lty=3, pch = 21, col = "blue")
legend("topright",
       legend = c(
         "Perfect Case: HG",
         "Perfect Case: BN",
         bquote(HG ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(BN ~ Imperfect ~ D[Delta] * "," ~ G[Lambda]),
         bquote(HG ~ Imperfect ~ D[delta] * "," ~ G[lambda]),
         bquote(BN ~ Imperfect ~ D[delta] * "," ~ G[lambda])
       ),
       col = c("black", "black", "red", "red", "blue", "blue"),
       lty = c(2, 3, 2, 3, 2, 3),     # match line types for each model
       pch = c(19, 21, 19, 21, 19, 21), # match point types for each model
       bty = "n", 
       cex = 0.8)
```

# Case Study: Salmonella in Pig Pens


## Application to Salmonella Surveillance on Pig Farms

Levels of \textit{Salmonella} bacteria on pig farms are regulated due to the high prevalence of infections and their significant contribution to human salmonellosis cases (Arnold, 2005). Farmed pigs are typically housed in pens of 30--100 individuals, and pooled faecal samples are tested routinely. Group testing is cost-efficient and improves detection because \textit{Salmonella} shedding in faeces is often intermittent. Furthermore, pooling can enhance the likelihood of detection due to collective sample representation. Farms are frequently ranked based on test outcomes (Arnold, 2005), and high prevalence may result in financial penalties. Therefore, understanding how different pooling strategies influence inferred prevalence is crucial for both epidemiological assessment and farm management.

Using a grouped-hypergeometric sampling framework, we extend the results in Arnold (2005) by incorporating biologically informed test-sensitivity models. We also show how Hellinger information can be applied to design efficient sampling strategies that improve the accuracy of prevalence estimates while controlling costs.

Now we assumes a single pen containing $N$ pigs, among which $T_X$ are infected. Faecal samples are collected and pooled for testing, with each pool composed of equal contributions from $\bar{N}$ individuals. Each individual contributes to at most one pool. The test sensitivity model, adapted from Cannon (2002) and applied by Arnold (2005), accounts for organism clustering in faecal matter, the pooling and homogenisation process, and the sub-sampling for PCR analysis. For a pool containing faecal material from $\bar{N}$ individuals, the individual-level test sensitivity $\delta$ is given by:

\begin{equation}
\label{pigdelta}
1 - \delta = \exp\left(-\frac{w_p C}{\bar{N}} \left(1 - \exp\left(-\frac{\widehat{\rho}}{w_p}\right)\right)\right) 
= \exp\left(-w_s C \left(1 - \exp\left(-\frac{\widehat{\rho}}{w_s \bar{N}}\right)\right)\right),
\end{equation}

\noindent where:
\begin{itemize}
  \item $w_s$ is the weight of faeces contributed by each individual,
  \item $w_p = w_s\bar{N}$ is the total pool weight,
  \item $w_a$ is the aliquot weight used for culturing,
  \item $C$ is the average number of clusters per gram,
  \item $\widehat{\rho} \approx \mathbb{E}(M) \rho w_a$ combines organism count per cluster, amplification probability, and aliquot weight.
\end{itemize}

For this application, $C = 7.3$ and $\widehat{\rho} = 0.55$ were estimated empirically in Arnold (2005). Assuming no false positives, the test specificity is set to $\lambda = 1$. The grouped-hypergeometric PMF incorporating imperfect test sensitivity is used to model the probability distribution of observed positive group tests. Based on this distribution, the Hellinger information are computed, providing a principled way to compare and optimize alternative sampling designs.



```{r Salmonela data structure}
N <- 60                      # Population size
barN <- c(20, 15, 10, 5, 1)  # Different group sizes
b <- rep(1,5)                # Number of groups
lambda <- 1                  # Item-level specificity 
wp <- 20                     # Total weight of the pooled sample
ws <- wp/barN                # Weight per item
rho_hat <- 0.55              # Empirical parameter for PCR
C <- 7.3                     # Average number of clusters per gram
Tx_values<-seq(0,60,by= 1)   # Range of contaminated items

# Sensitivity function (Equation 19) 
calculate_delta_ws <- function(barN, ws, rho_hat, C) {
  1 -  exp(-ws * C * (1 - exp(-rho_hat / (ws*barN))))
}
# Calculate item-level sensitivity (delta)
item_level_sensitivities_case2 <- calculate_delta_ws(barN, ws, rho_hat, C)
plot(barN,item_level_sensitivities_case2,xlab = expression(bar(N)),ylab=expression(delta),main=expression(delta * " against " * bar(N) ))
```

Now we can estimate HI for $T_X$ values under different item-level sensitivity.

```{r HI delta pig case}
# Define helper safely extracting FI_Tx
extract_FI <- function(x) {
  if (is.list(x) && "FI_Tx" %in% names(x)) {
    return(x$FI_Tx)
  } else {
    return(0)
  }
}

fisher_results_HG_PMF <- lapply(seq_along(barN), function(i) {
  n <- barN[i]
  b_i <- b[i]
  ws_i <- ws[i]
  delta <- calculate_delta_ws(
    barN = n,
    ws = ws_i,
    rho_hat = rho_hat,
    C = C
  )
  lambda <- 1
  fisher_info <- lapply(Tx_values, function(Tx) {
    info_hg_tx_imperfect(
      N = N,
      barN = n,
      Tx = Tx,
      b = b_i,
      delta = delta,
      lambda = lambda,
      method = "PMF-HI",
      type = "item",
      verbose = FALSE
    )
  })
  HI <- sapply(fisher_info, extract_FI)
  list(
    barN = n,
    b = b_i,
    ws = ws_i,
    delta = delta,
    lambda = lambda,
    HI = HI
  )
})


```

Now we plot the trends of HI for given $T_X$ in below Figure. The Figure shows Hellinger information for the grouped-hypergeometric distribution as a function of the number of infected individuals ($T_X$), assuming a pen size of $N = 60$ and a total sample weight of 20 grams. The sample is treated as a single group ($b = 1$), with different pooling sizes ($\bar{N} = 20, 10, 5, 4, 2, 1$) representing how many individuals contributed to the pooled sample. Red vertical lines indicate reference prevalence levels: the average reported ($T_X = 15$), a low ($T_X = 9$), and a high ($T_X = 51$) contamination level. Sensitivity is derived from a fecal concentration model shown above; specificity is assumed perfect ($\lambda = 1$).

```{r HI delta pig case Fig}
colors <- c("black", "blue", "green", "purple", "orange")
range_FI <- range(fisher_results_HG_PMF[[1]]$HI,
                  fisher_results_HG_PMF[[2]]$HI,
                  fisher_results_HG_PMF[[3]]$HI,
                  fisher_results_HG_PMF[[4]]$HI,
                  fisher_results_HG_PMF[[5]]$HI)

plot(Tx_values[c(2:60)], fisher_results_HG_PMF[[1]]$HI[c(2:60)], type = "l",
  xlab = expression(paste("Number of contaminated items ", T[x])), ylab = "Hellinger information (HI)",
  main = "HI: Grouped-hypergeometric model",
  col = colors[1], lwd = 2,ylim=range_FI,xaxt = "n",yaxt="n")
axis(side = 1, at = seq(0, 60, by = 5))
axis(side = 2, at = seq(0, range_FI[2], by = 0.005))
for (i in 2:length(barN)) {
  lines(Tx_values[c(2:60)], fisher_results_HG_PMF[[i]]$HI[c(2:60)], type = "l", col = colors[i], lwd = 2, lty = i)
}
legend_labels <- lapply(seq_along(barN), function(i) {
  substitute(b == x * ~ ", " ~ bar(N) == y * ", " ~ delta == z, list(x = b[i], y =  barN[i], z =  round(item_level_sensitivities_case2[i],2)))
})
legend(
  20,0.04, legend = legend_labels,
  col = colors, lty = 1:length(barN), cex = 1, bty = "n"
)
abline(v = c(9, 15, 51), col = c(2, 2, 2), lwd = 2)
line_labels <- c(expression(T[x] == 0.15 * N ~ " = 9"),
                 expression(T[x] == 0.25 * N ~ " = 15"),
                 expression(T[x] == 0.85 * N ~ " = 51"))
text(x = c(9, 15, 51), y = par("usr")[4]*0.5, labels = line_labels, pos = 3, col = 2, cex = 0.7,srt = 90, adj = 0)  

```



## Design exploration: **Grid search over** $b$ and $\bar{n}$

Here we wish to explore an optimal design based on HI. For this purporse, we estimates HI for all of sort combination of $b$ and $\bar{N}$ for a given $T_X=15$. We use the pig data as well where sensitivity is a function of $\bar{N}$. 

First we create a grid of $b$, $\bar{N}$, $w_s$ and $\delta$ and then we will estimate HI for $T_X=15$. Later we will use the same grid to estimate HI assuming approximate BN model. Then we will compare the HI heatmap to explore a better sampling design based on the HI.


```{r grid}
N <- 60 # Population size
max_group_size <- 20
max_groups <- 20
wn <- 20         # a total weight of faecal matter tested
lambda <- 1      # Specificity for item-level
rho_hat <- 0.55  # Empirical parameter for PCR
C <- 7.3  # Average number of clusters per gram
Tx_values <- c(15)  
ws <- c()
b_store <- c()
barN_store <- c()
for (b in 1:max_groups) {
  for (N_bar in 1:max_group_size) {
    if (b * N_bar <= N) {
      b_store <- c(b_store, b)
      barN_store <- c(barN_store, N_bar)
      ws <- c(ws, wn / (b * N_bar))
    }
  }
}
combinations <- data.frame(b = b_store, barN = barN_store, ws = ws)
combinations$delta <- calculate_delta_ws(combinations$barN, ws=combinations$ws, rho_hat, C)
head(combinations)
```

Now we calculate HI for each combination based on HG model.

```{r HI HG Pig Case Heatmap}
fisher_results_HG_PMF <- lapply(1:dim(combinations)[1], function(i) {
  n <- combinations$barN[i]
  b_i <- combinations$b[i]
  ws_i <- combinations$ws[i]
  delta <- combinations$delta[i]
  lambda <- 1
  fisher_info <- lapply(Tx_values, function(Tx) {
    info_hg_tx_imperfect(
      N = N, barN = n, Tx = Tx, b = b_i,
      delta = delta, lambda = lambda,
      method = "PMF-HI", type = "item", verbose = FALSE
    )
  })
  HI <- sapply(fisher_info, extract_FI)
  list(
    barN = n,
    b = b_i,
    ws = ws_i,
    delta = delta,
    lambda = lambda,
    HI = HI
  )
})

```


In similar way we calculate HI for each combination based on approximate BN model.

```{r HI BN Pig Case Heatmap}
fisher_results_BN_PMF <- lapply(1:dim(combinations)[1], function(i) {
  n <- combinations$barN[i]
  b_i <- combinations$b[i]  # Ensure correct mapping of b to barN
  ws_i <- combinations$ws[i]
  delta <- combinations$delta[i]
  lambda <- 1
  fisher_info <- sapply(Tx_values, function(Tx) {
    info_bn_tx_imperfect(
      N = N, barN = n, Tx = Tx, b = b_i,
      delta = delta, lambda = lambda,
      method = "PMF-HI", type = "item"
    )
  })
  list(barN = n, b=b_i,ws=ws_i,delta=delta,lambda=lambda,HI = fisher_info)
})
HI_BN_Tx_15 <- rep(NA, dim(combinations)[1])
for (i in 1: dim(combinations)[1]) {
  HI_BN_Tx_15[i] <- fisher_results_BN_PMF[[i]]$HI
}
combinations$HI_BN_Tx_15 <- HI_BN_Tx_15
combinations_BN_Tx_15 <- combinations[,c("b","barN","HI_BN_Tx_15")]
head(combinations_BN_Tx_15)
```

We created heatmap for both cases.
```{r heatmap Tx 15}
# Recompute HI for each combination (grid search) with all assignments
fisher_results_HG_PMF <- lapply(1:dim(combinations)[1], function(i) {
  n <- combinations$barN[i]
  b_i <- combinations$b[i]
  ws_i <- combinations$ws[i]
  delta <- combinations$delta[i]
  lambda <- 1
  fisher_info <- lapply(Tx_values, function(Tx) {
    info_hg_tx_imperfect(
      N = N,
      barN = n,
      Tx = Tx,
      b = b_i,
      delta = delta,
      lambda = lambda,
      method = "PMF-HI",
      type = "item",
      verbose = FALSE
    )
  })
  HI <- sapply(fisher_info, extract_FI)
  list(
    barN = n,
    b = b_i,
    ws = ws_i,
    delta = delta,
    lambda = lambda,
    HI = HI
  )
})

# Extract HI for Tx = 15
HI_HG_Tx_15 <- vapply(fisher_results_HG_PMF, function(res) {
  # Defensive indexing in case HI vector shorter
  if (length(res$HI) < 16) return(0)
  res$HI[16]
}, numeric(1))

# Store results in combinations data frame
combinations$HI_HG_Tx_15 <- HI_HG_Tx_15

# Make table for plotting
combinations_HG_Tx_15 <- combinations[, c("b", "barN", "HI_HG_Tx_15")]


min_value <- min(combinations_HG_Tx_15$HI_HG_Tx_15, combinations_BN_Tx_15$HI_BN_Tx_15, na.rm = TRUE)
max_value <- max(combinations_HG_Tx_15$HI_HG_Tx_15, combinations_BN_Tx_15$HI_BN_Tx_15, na.rm = TRUE)

Tx_15_HG <- ggplot(combinations_HG_Tx_15, aes(x = barN, y = b, fill = HI_HG_Tx_15)) +
  geom_tile() + 
  scale_fill_viridis_c(option = "mako", direction = -1,name = NULL, n.breaks = 10,limits = c(min_value, max_value),
                       guide = guide_colourbar(barheight = unit(6, "cm"))) + 
  labs(x = expression("Group size (" ~ bar(N) ~ ")"),
       y = expression("Number of groups (" ~ b ~ ")"),
       title = "Grouped-hypergeometric model",
       subtitle = expression("Hellinger information–HG, " ~ T[x] ~ "= 15")) +
  theme_minimal(base_size = 14) +     
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))


Tx_15_BN <- ggplot(combinations_BN_Tx_15, aes(x = barN, y = b, fill = HI_BN_Tx_15)) +
  geom_tile() +                       
  scale_fill_viridis_c(option = "mako", direction = -1,name = NULL, n.breaks = 10,limits = c(min_value, max_value),
                       guide = guide_colourbar(barheight = unit(6, "cm"))) + 
  labs(x = expression("Group size (" ~ bar(N) ~ ")"),
       y = expression("Number of groups (" ~ b ~ ")"),
       title = "Grouped-binomial model",
       subtitle = expression("Hellinger information–BN, " ~ T[x] ~ "= 15")) +
  theme_minimal(base_size = 14) +       # Clean theme
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

ggarrange(Tx_15_HG,Tx_15_BN,ncol = 2,nrow = 1)
```

# Session Info & References

```{r Session}
sessionInfo()
```

# References

- Barnes, B., Parsa, M., Das, S., & Clark, R. (2025). *Hypergeometric and Binomial Group Sampling with Sensitivity and Specificity*. Submitted to *Communications in Statistics -- Theory and Methods*.
- Arnold, M. E., Cook, A., & Davies, R. (2005). A modelling approach to estimate the sensitivity of pooled faecal samples for isolation of Salmonella in pigs. *Journal of the Royal Society Interface*, 2(4), 365–372.
- Theobald, C. M., & Davie, A. M. (2014). Group testing, the pooled hypergeometric distribution, and estimating the number of defectives in small populations. *Communications in Statistics -- Theory and Methods*, 43(14), 3019–3026.
